mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-02 Draft1.Rmd
to do

@@

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
However, so far, R expressions such as `pbinom(k, N, p)` are printed as-is; 
prettier mathematical expressions such as $P_\mathrm{Bi}(X \le k; N, p)$ require
Latex commands like `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for
minimalistic examples, these commands are tedious to type in and hard to read in
the original code. Since recently [@R, version 4.2], manual pages include
support for mathematical expressions [@Sarkar2022], which already is a big
improvement. Rules for the translation of R's built-in language elements to
their mathematical representation are lacking.

@@

@@

Here we present a second less trivial example where Prolog needs to find out the name of
the integration variable for `sin`. For that purpose, rolog provides a predicate `r_eval/2` that calls R from Prolog. Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.

@@

Note that the Prolog end, the handler for `integrate/3` is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, `integrate(sin, lower=0L, upper=2L * pi)` would not print the desired
result.

@@

# Plus Minus
catmath(x %+-% y)
catmathjax(x %+-% y)

@@

# %*%
catmath(x %*% y)
catmathjax(x %*% y)

# %.%
catmath(x %.% y)
catmathjax(x %.% y)

@@

# %/%: integer division
catmath(x %/% y)
catmathjax(x %/% y)
```

@@

catmath(choose(k=K, n=N))
catmathjax(choose(k=K, n=N))

@@

@@

@@

catmath(integrate(x))
catmathjax(integrate(x))

@@

@@

@@

This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them more smoothly. Building on current features of R\ and existing packages for displaying formulas in R\ (e.g. Murrell and Ihaka [-@murrell2000], Allaire et al. [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules, via the \pkg{rolog}, and translates the R\ expressions to MathML. Then, two possible solutions for typesetting are available: LaTeX and MathJax to export R\ Markdown documents in different formats or ShinyApp webpages. The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.
In its current version \pkg{mathml} has some limitations. For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation. However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations. Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.

>>>>

#### To do

* paste(x, y, z)
* Spacing
* Comma separated lists
* Ellipsis
* Radicals
* Subsets
* Greek letters in MathJax
* Typeface
* Stylechanges
* Phantoms

@@

The R extension of the markdown language [@Xie2020] enables reproducible
statistical reports with nice typesetting in HTML, Microsoft Word, and Latex.
However, so far, R expressions such as `pbinom(k, N, p)` are printed as-is; 
prettier mathematical expressions such as $P_\mathrm{Bi}(X \le k; N, p)$ require
Latex commands like `P_\mathrm{Bi}\left(X \le k; N, p\right)`. Except for
minimalistic examples, these commands are tedious to type in and hard to read in
the original code. Since recently [@R, version 4.2], manual pages include
support for mathematical expressions [@Sarkar2022], which already is a big
improvement. However, rules for the translation of R's built-in language elements to
their mathematical representation are still lacking.

@@

## Prolog: the \pkg{rolog} package

@@

Here we present a second less trivial example where Prolog needs to find out the name of
the integration variable for `sin`. For that purpose, \pkg{rolog} provides a predicate `r_eval/2` that calls R from Prolog. Here, the predicate is used for the  R\ function `formalArgs(args(sin))`, which returns the name of the function argument of `sin`, that is, `x`.

@@

Note that, on the Prolog end, the handler for `integrate/3` is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, `integrate(sin, lower=0L, upper=2L * pi)`, would not print the desired
result.

@@

catmath(x %+-% y)
catmathjax(x %+-% y)

@@

catmath(x %*% y)
catmathjax(x %*% y)

catmath(x %.% y)
catmathjax(x %.% y)

@@

catmath(x %/% y)
catmathjax(x %/% y)
```

## Subscripts and Superscripts

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(a["x"])
catmathjax(a["x"])

catmath(a^x)
catmathjax(a^x)

catmath(x["b"]^"a")
catmathjax(x["b"]^"a")

@@

@@

## Radicals

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}

catmath(sqrt(x))
catmathjax(sqrt(x))

@@

## Accents

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}
catmath(hat(x))
catmathjax(hat(x))

catmath(hat(xy))
catmathjax(hat(xy))

catmath(mean(x))
catmathjax(mean(x))

catmath(mean(xy))
catmathjax(mean(xy))

## Symbolic names

```{r, results="asis", echo=TRUE,  eval=knitr::is_html_output()}

catmath(Alpha-Omega)
catmathjax(Alpha-Omega)

catmath(alpha-omega)
catmathjax(alpha-omega)

catmath(Inf)
catmathjax(Inf)

@@

catmath(integrate(sin, 0L, 2L*pi))
catmathjax(integrate(sin, 0L, 2L*pi))

@@

## Grouping

```{r, results='asis', echo=TRUE,  eval=knitr::is_html_output()}
catmath(x^y + z)
catmathjax(x^y + z)

catmath(x^(y + z))
catmathjax(x^(y + z))

# without parenthesis
mathml(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))
mathjax(canonical(call("^", as.symbol("x"), call("+", as.symbol("y"), as.symbol("z")))))

# shortcut
catmath(x^{y + z})
catmathjax(x^{y + z})

@@

## Matrix and Vectors

Since also Matrices and Vectors are important elements in R\, we decided to implement them and render them graphically in the following ways.

vec <- 1:3
mathml(canonical(vec))
mathml(vec)
mathjax(canonical(vec))
mathjax(vec)

mat <- matrix(data = c(2L, 3L, 4L, 5L), nrow = 2, ncol = 2)
mathml(canonical(mat))
mathjax(canonical(mat))

@@

This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them smoothly. Building on current features of R\ and existing packages for displaying formulas in R\ (e.g. Murrell and Ihaka [-@murrell2000], Allaire et al. [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML. Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax.

The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.

In its current version \pkg{mathml} has some limitations. For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation. However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations. Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.
